---
title: "Deployment report"
format:
  html:
    theme: [default, report.scss]
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-location: left
    fig-align: center
    page-layout: article
params:
  bundle: ""
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(birdnetprocess)
library(ggplot2)

b <- readRDS(params$bundle)
cfg <- b$cfg
st <- b$stats

knitr::opts_chunk$set(
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
  dpi = 150,
  fig.path = "figures/",
  fig.retina = 2,
  out.width = "100%"
)

# Compact numbers read faster in a tile than exact ones; the exact count is in
# the tables below for anyone who needs it.
compact <- function(n) {
  if (is.na(n)) return("—")
  if (n >= 1e6) return(paste0(round(n / 1e6, 1), "M"))
  if (n >= 1e4) return(paste0(round(n / 1e3), "K"))
  format(n, big.mark = ",")
}

tile <- function(label, value, note = NULL) {
  sprintf(
    '<div class="bn-tile"><div class="bn-tile-label">%s</div>
     <div class="bn-tile-value">%s</div>%s</div>',
    label, value,
    if (is.null(note)) "" else sprintf('<div class="bn-tile-note">%s</div>', note)
  )
}

fmt_date <- function(x) if (is.na(x)) "—" else format(x, "%d %b %Y")

`%||%` <- function(x, y) if (is.null(x)) y else x

# Emit HTML as an explicit raw block. Without the fence, pandoc parses the
# indented continuation lines as markdown and drops the closing tags.
raw_html <- function(...) cat("\n```{=html}\n", paste0(...), "\n```\n", sep = "")

# Figure height scales with the number of species rows, or a tall plot gets
# squashed into unreadable slivers. Computed here so the chunks below can use
# it as a chunk option.
n_species_shown <- function(p) {
  if (is.null(p) || is.null(p$data)) return(10)
  length(unique(p$data[["Common Name"]]))
}
activity_height <- max(4, min(22, 2 + 0.24 * n_species_shown(b$plots$activity)))
counts_height <- max(4, min(12, 1.5 + 0.28 * n_species_shown(b$plots$counts)))

# Wide tables scroll inside their own box so the page never scrolls sideways.
html_table_str <- function(df, align_right = character(0)) {
  body <- if (requireNamespace("gt", quietly = TRUE)) {
    g <- gt::gt(df)
    if (length(align_right) > 0) {
      g <- gt::cols_align(g, align = "right", columns = align_right)
    }
    g <- gt::tab_options(
      g,
      table.font.size = "0.85rem",
      table.background.color = "transparent",
      table.border.top.style = "none",
      table_body.hlines.color = "#e1e0d9",
      column_labels.border.bottom.color = "#c3c2b7"
    )
    as.character(gt::as_raw_html(g))
  } else {
    paste(knitr::kable(df, format = "html", row.names = FALSE,
                       format.args = list(big.mark = ",")), collapse = "\n")
  }
  paste0('<div class="bn-scroll">', body, '</div>')
}

html_table <- function(df, align_right = character(0)) {
  raw_html(html_table_str(df, align_right))
}

# Housekeeping the report has to state but nobody reads top-to-bottom: what was
# filtered out and why. Folded shut so it costs one line of the page, and is
# still there for anyone who asks where a species went.
details <- function(summary, ...) {
  sprintf('<details class="bn-details"><summary>%s</summary>%s</details>',
          summary, paste0(...))
}
```

```{r header}
#| output: asis
span <- if (identical(fmt_date(st$first), fmt_date(st$last))) {
  fmt_date(st$first)
} else {
  paste(fmt_date(st$first), "–", fmt_date(st$last))
}

raw_html(sprintf(
  '<div class="bn-header">
<h1 class="bn-title">%s</h1>
<p class="bn-subtitle">%s · %s</p>
<dl class="bn-meta">
<div><dt>Detector</dt><dd>%s</dd></div>
<div><dt>Threshold</dt><dd>confidence ≥ %s</dd></div>
<div><dt>Coordinates</dt><dd>%.4f, %.4f</dd></div>
<div><dt>Time zone</dt><dd>%s</dd></div>
<div><dt>Generated</dt><dd>%s</dd></div>
</dl>
</div>',
  cfg$name,
  cfg$location$place %||% "",
  span,
  paste0(cfg$detector$name,
         if (!is.null(cfg$detector$version)) paste0(" v", cfg$detector$version) else ""),
  st$confidence,
  cfg$location$latitude, cfg$location$longitude,
  cfg$location$tz,
  format(b$generated, "%d %b %Y, %H:%M")
))
```

```{r tiles}
#| output: asis
raw_html(
  '<div class="bn-tiles">',
  tile("Sites", st$n_sites),
  tile("Species", compact(st$n_species), "distinct common names"),
  tile("Detections", compact(st$n_detections),
       sprintf("at confidence ≥ %s", st$confidence)),
  tile("Recordings", compact(st$n_recordings), "source audio files"),
  '</div>'
)
```

## What was recorded

```{r per-site}
#| output: asis
tbl <- b$per_site
tbl$span <- ifelse(
  format(tbl$first, "%Y-%m-%d") == format(tbl$last, "%Y-%m-%d"),
  format(tbl$first, "%d %b %Y"),
  paste(format(tbl$first, "%d %b"), "–", format(tbl$last, "%d %b %Y"))
)
out <- data.frame(
  Site = tbl$Site,
  Dates = tbl$span,
  Recordings = format(tbl$recordings, big.mark = ","),
  Detections = format(tbl$detections, big.mark = ","),
  Species = tbl$species,
  check.names = FALSE
)

html_table(out, align_right = c("Recordings", "Detections", "Species"))
```

::: {.bn-note}
Counts are detections, not individual animals: one bird calling repeatedly
produces many rows. Read them as calling activity, not abundance.
:::

```{r silent-sites}
#| output: asis
if (length(b$silent_sites) > 0) {
  cat("::: {.bn-note .bn-warn}\n")
  cat("**", length(b$silent_sites), " of ", st$n_sites + length(b$silent_sites),
      " sites contributed nothing to this report.** They recorded, and the ",
      "detector scored them, but no detection reached the reporting threshold, ",
      "so they appear in no table or figure below:\n\n", sep = "")
  cat(paste0("- ", b$silent_sites, collapse = "\n"))
  cat("\n\nA whole site scoring far below the others usually means a hardware ",
      "fault rather than an absence of birds — check the recordings' audio ",
      "levels before concluding the site was quiet.\n", sep = "")
  cat(":::\n\n")
}
```

## Which species, and how much

```{r counts}
#| fig-width: 10
#| fig-height: !expr counts_height
#| fig-alt: "Bar chart of detection counts per species, most detected first."
if (!is.null(b$plots$counts)) b$plots$counts
```

## When they called

The diel plot folds every recorded day onto one 24-hour clock, so a species'
row shows the time of day it calls rather than which day it happened to be
heard. Each row is scaled to its own peak — rows show *timing*, not relative
abundance between species. Grey bands are night.

```{r activity}
#| fig-width: 12
#| fig-height: !expr activity_height
#| fig-alt: "Heatmap of calling activity by time of day, one row per species."
if (!is.null(b$plots$activity)) b$plots$activity
```

```{r trends}
#| fig-width: 10
#| fig-height: 6
#| fig-alt: "Detections per hour over the deployment for the most-detected species."
if (!is.null(b$plots$trends)) b$plots$trends
```

## How much to trust it

```{r confidence}
#| fig-width: 9
#| fig-height: 7
#| fig-alt: "Boxplots of detection confidence for the most-detected species."
if (!is.null(b$plots$confidence)) b$plots$confidence
```

::: {.bn-note}
A species whose confidence sits just above the threshold deserves a listen
before it goes in a results table. Detector confidence is not a probability of
correctness, and it is not comparable between species.
:::

```{r clips}
#| output: asis
if (b$n_clips > 0) {
  scope <- if (b$clip_species < st$n_species) {
    sprintf("the %d most-detected species", b$clip_species)
  } else {
    "every species"
  }
  cat(sprintf(
"### Listen for yourself

The %d highest-confidence clips of %s sit in `clips/` beside this report, with
a listening sheet at [`clips/index.html`](clips/index.html): play each clip,
mark whether it is the bird, and send back the CSV it exports.
[`%s`](%s) is the same folder packed to email — the sheet needs the audio
alongside it, so send the archive rather than the page.

These are each species' best clips, which makes this the friendliest test the
detector can be given. It settles which species are wrong; it cannot vouch for
the quieter detections behind the ones you hear%s.

", b$n_clips, scope, b$clips_zip, b$clips_zip,
   if (b$clip_species < st$n_species) {
     sprintf(", nor for the %d species with no clips here",
             st$n_species - b$clip_species)
   } else {
     ""
   }))
}
```

```{r thresholds}
#| output: asis
if (isTRUE(st$per_species)) {
  cat(sprintf(
    "\n### Thresholds\n\n%d of the %d species reported here were thresholded at their own calibrated operating point; the rest use the deployment default of %s.\n\n",
    st$n_calibrated, st$n_species, st$confidence
  ))
} else {
  cat("\n### Thresholds\n\n")
  # The config's own note leads when there is one; a generic sentence in front
  # of it just says the same thing twice.
  if (is.null(b$threshold_note)) {
    cat(sprintf("Every species here is cut at the same confidence, %s.\n\n",
                st$confidence))
  }
}

# The note explains where the number came from, which matters just as much
# when there is only one of them.
if (!is.null(b$threshold_note)) {
  cat("::: {.bn-note}\n", b$threshold_note, "\n:::\n\n", sep = "")
}

if (length(b$flagged) > 0) {
  cat("::: {.bn-note .bn-warn}\n")
  cat("**Not validated.** Evaluation on labelled audio did not establish a ",
      "reliable operating point for these species, so their detections are ",
      "shown for review rather than as findings, and should be listened to ",
      "before being counted:\n\n", sep = "")
  cat(paste0("- ", b$flagged, collapse = "\n"))
  cat("\n\nThat is a statement about the labelled soundscape the detector was ",
      "scored on, not a measurement made here — a species can fail there ",
      "simply by being too rare in it to assess.\n", sep = "")
  cat(":::\n\n")
}
```

## Full species list

```{r species-table}
#| output: asis
out <- b$species_table
if ("basis" %in% names(out)) {
  out$basis <- c(calibrated = "calibrated", unvalidated = "NOT VALIDATED",
                 default = "default")[out$basis]
}
names(out) <- c("Species", "Detections", "Sites", "Median confidence",
                "Max confidence", "Threshold", "Basis")[seq_along(out)]
# Under a single threshold the last two columns are dropped upstream, so align
# only what is actually here.
html_table(out, align_right = intersect(
  c("Detections", "Sites", "Median confidence", "Max confidence", "Threshold"),
  names(out)
))
```

## What was filtered out

```{r filtering}
#| output: asis
ex <- b$excluded
if (!is.null(ex) && nrow(ex) > 0) {
  lines <- sprintf(
    "- **%s** — %s detections removed%s",
    ex$species, format(ex$detections, big.mark = ","),
    ifelse(nzchar(ex$reason), paste0(". ", ex$reason), "")
  )
  cat("Species named in the config as not occurring here, removed before any ",
      "figure or table below was computed:\n\n", sep = "")
  cat(paste(lines, collapse = "\n"), "\n\n", sep = "")
}

rr <- b$range_report
if (!is.null(rr)) {
  removed <- rr[grepl("^removed", rr$status), , drop = FALSE]
  unassessed <- sum(grepl("^unassessed", rr$status))
  head_line <- sprintf(
    "Range filter: %d of %d species dropped as geographically implausible here; %d had no range map and were kept.",
    nrow(removed), nrow(rr), unassessed
  )
  if (nrow(removed) > 0) {
    tbl <- removed[order(-removed$n_detections),
                   c("species", "n_detections", "max_confidence", "status")]
    names(tbl) <- c("Species", "Detections", "Max confidence", "Outcome")
    raw_html(details(
      paste0(head_line, " <span class=\"bn-more\">Show the list</span>"),
      '<p class="bn-note">Modelled abundance at these coordinates is zero, so a ',
      'detection here is a misidentification of something that does occur. A ',
      'species genuinely present but absent from the range model would be ',
      'wrongly dropped — check this list if you expected something that is not ',
      'in the report.</p>',
      html_table_str(tbl, align_right = c("Detections", "Max confidence"))
    ))
  } else {
    cat(head_line, "\n\n")
  }
}
```

## How this report was made

```{r provenance}
#| output: asis
model <- cfg$detector$model
model_note <- if (is.null(model)) {
  "stock model (no custom classifier configured)"
} else if (file.exists(model)) {
  sprintf("`%s` (md5 `%s`)", basename(model),
          substr(tools::md5sum(model), 1, 12))
} else {
  sprintf("`%s` (file not found at report time)", basename(model))
}

cat(sprintf(
  "- **Config**: `%s`\n- **Detector**: %s\n- **Model file**: %s\n- **Sites**: %s\n- **birdnetprocess**: %s\n- **R**: %s\n",
  cfg$path,
  paste0(cfg$detector$name,
         if (!is.null(cfg$detector$version)) paste0(" v", cfg$detector$version) else ""),
  model_note,
  paste(vapply(cfg$sites, function(s) s$id, character(1)), collapse = ", "),
  as.character(utils::packageVersion("birdnetprocess")),
  R.version.string
))

if (length(b$failures) > 0) {
  cat('\n::: {.bn-note .bn-warn}\n')
  cat("Some figures could not be built and are missing above:\n\n")
  cat(paste0("- ", b$failures, collapse = "\n"))
  cat("\n:::\n")
}
```
